# The `smfsb` R package

## Introduction

In the [previous post](https://darrenjw.wordpress.com/2018/12/19/stochastic-modelling-for-systems-biology-third-edition/) I gave a brief introduction to the third edition of my textbook, [Stochastic modelling for systems biology](https://github.com/darrenjw/smfsb). The algorithms described in the book are illustrated by implementations in R. These implementations are collected together in an R package on [CRAN](https://cran.r-project.org/) called `smfsb`. This post will provide a brief introduction to the package and its capabilities.

## Installation

Since the package is on CRAN, the simplest way to install it is to enter
```{r eval=FALSE}
install.packages("smfsb")
```
at the R command prompt. This will install the latest version that is on CRAN. Once installed, the package can be loaded with
```{r message=FALSE}
library(smfsb)
```
The package is well-documented, so further information can be obtained with the usual R mechanisms, such as
```{r eval=FALSE}
vignette(package="smfsb")
vignette("smfsb")
help(package="smfsb")
?StepGillespie
example(StepCLE1D)
```


## A brief tutorial

The vignette gives a quick introduction the the library, which I don't need to repeat verbatim here. If you are new to the package, I recommend working through that before continuing. Here I'll concentrate on some of the new features associated with the third edition.




Some text

Some code:

```{r}
x=rnorm(5)
y=x*x
y
```

Some more text

```{r}
plot(x,y)
```

blah


## Source

This blog post started out as an [RMarkdown](https://rmarkdown.rstudio.com/) document, the source of which can be found [here]() LINK.

#### eof

